---
author: "Alexander Gibson"
categories:
date: "2024-11-27"
draft: false
excerpt: How I became interested in meta-research and how to build the community by engaging students.
layout: single
subtitle: How I see the future of meta-research
title: Optimism in Meta-Research
auto_ids: false
---
I got to experience my first insight of ‘meta-research’ at the 2023 [AIMOS conference](https://aimos.community/). Before this, I had almost no knowledge about meta-research. It sparked an interest in me for integrity and good research practices. Often meta-research can appear pessimistic or cynical of broader research, showcasing the flaws and faults, which made it refreshing to see a dedicated session for ‘some good news’ at AIMOS 2024. Four talks presented by [David Mellor](https://www.cos.io/team/david-mellor), [Tammy Hoffman](https://bond.edu.au/profile/tammy-hoffmann), [Ivan Oransky](https://retractionwatch.com/meet-the-retraction-watch-staff/about/) and [Anita Bandrowski](https://profiles.ucsd.edu/anita.bandrowski) gave light to the progress and positive effects meta-research and open science is having on the wider research. I found it very uplifting. The steady growth of articles relating to 'meta-research' is even more encouraging (Figure 1). Even organisations like the [Center for Open Science](https://www.cos.io/), [Research on Research Institute](https://researchonresearch.org/), [AIMOS](https://aimos.community/) or [Metascience](https://metascience.info/) previously did not exist, yet now they do. They are only a few examples of direct evidence for the progress being made. 

<br>

```{r include=FALSE}
# load libraries
library(rentrez)
library(ggplot2)
library(tidyverse)
library(glue)

# set the time frame for the search
year <- 2000:2024

# combine different search terms and dates to be extracted from rentrez
meta <- glue("meta-research AND {year}[PDAT]")

# get data of published articles from the terms from pubmed 
df <- tibble(year = year,
                    meta = meta) %>% 
                      mutate(meta = map_dbl(meta, ~entrez_search(db = "pubmed", term = .x)$count))


# plot the figure
df %>% 
  select(year, meta) %>% 
  filter(year<2024) %>% 
  ggplot(aes(x = year)) +
  geom_line(aes(y = meta),linewidth = 2, show.legend = FALSE) +
  theme_void() +
  labs(y = "",
       x = "") +
  theme(panel.border = element_rect(linewidth = 4, fill = NA))

# save the figure to figures folder
ggsave(filename = "featured.jpg", width = 8, height = 6, dpi = 300)
```

```{r echo=FALSE}
# plot the figure
df %>% 
  select(year, meta) %>% 
  filter(year < 2024) %>% 
  ggplot(aes(x = year)) +
  geom_line(aes(y = meta),linewidth = 1, show.legend = FALSE) +
  theme_classic() +
  labs(y = "Number of Articles",
       x = "Year") +
  scale_x_continuous(limits = c(2000, 2024), breaks = c(2000,2008,2015,2023))+
  scale_y_continuous(limits = c(0,250))+
  labs(caption = "Figure 1: Published articles per year between 2000 and 2023 in PubMed containing
'meta-research' in the title or abstract.")+
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0, size = 12),
        axis.title.x = element_text(size = 14),  # Increase x-axis title size
        axis.title.y = element_text(size = 14),  # Increase y-axis title size
        axis.text.x = element_text(size = 12),   # Increase x-axis label size
        axis.text.y = element_text(size = 12))   # Increase y-axis label size
```

As an early PhD student, I was finishing high school only six years ago. I’m positioned with potentially a few decades of research to come. This fills me with optimism looking ahead for new possibilities. Especially, with students much like myself that are becoming more and more interested in meta-research. New researchers will help to expand and grow the discipline, bringing new ideas and future outcomes. [Joan Leach](https://cpas.anu.edu.au/people/professor-joan-leach), a plenary speaker at AIMOS 2024 put it perfectly, that “undergraduates need touch points with meta-research from day one”. Involving students early, opens opportunities and pathways to engage with meta-research. It worked for me. 


But before a student can decide if they're interested, they must first know about meta-research. This means getting meta-research in front of new students. Simply, an opportunity arose for me to help with the set up / audio-visuals at the 2023 AIMOS conference. My curiosity and interest were brought about from talks by [Eugenie Reich](https://eugeniereichlaw.com/), [Danielle Oste](https://www.linkedin.com/in/danielle-oste-35194729b/?originalSubdomain=au) and [Carl Bergstrom](https://ctbergstrom.com/index.html). A particular talk by [Laure Wynants](https://www.maastrichtuniversity.nl/l-wynants) on clinical prediction models was one I found incredibly insightful, leaving my mind conjuring a plethora of potential research questions and ideas. Part of my interest sparked then, has developed into my current PhD. I believe involving and disseminating topics of meta-research to students will be one of the key drivers in expanding the community of future researchers. 

**How do we involve new students?**

Most obvious is to increase the number of honours, MPhil and PhD projects that are focused on meta-research (easier said than done), providing the opportunity for students to pursue their interests. However, this would only be possible for already established researchers who can provide these resources to students. This further assumes the students are already interested in these topics. If not, lectures, workshops or conferences would be effective tools in engaging a larger number of students to spark that interest --- how are you engaging with students? Increasing the availability and quantity of these events would again, be beneficial. Starting a new form of content, whether that be a blog, podcast or social media post will make meta-research more accessible to a wider audience. A month ago the 'Curiosity chronicles' didn't exist, yet here you are reading it. 

Looking to the future, I remain excited about the potential meta-research will have on shaping policy and practice. By involving students early and providing the resources to explore this area of research, it will help to build the community. Whether it is academically specific or through informal content, these touch points with students are essential. The field of meta-research is just beginning and the potential for impact is great.